---
title: "Shapley Additive Global Importance (SAGE)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
	collapse = TRUE,
	comment = "#>",
	fig.width = 8,
	fig.height = 6
)
set.seed(123)
options("xplain.progress" = interactive())
```

```{r setup}
library(xplainfi)
library(mlr3)
library(mlr3learners)
library(ggplot2)
```

## Introduction

Shapley Additive Global Importance (SAGE) is a feature importance method based on cooperative game theory. It uses Shapley values to distribute the model's total prediction performance among features. Unlike perturbation-based methods (PFI/CFI) that measure performance degradation when features are perturbed, SAGE measures each feature's contribution through marginalization.

A key property of SAGE is that it provides a complete decomposition: the sum of all SAGE values equals the difference between the model's performance and the performance when all features are marginalized.

`xplainfi` provides two implementations:

- **MarginalSAGE**: Marginalizes features independently (standard SAGE implementation)
- **ConditionalSAGE**: Marginalizes features using conditional sampling

**Note on interpretation**: SAGE's theoretical properties and interpretation differ from perturbation-based methods. While PFI/CFI have clearer interpretations in terms of predictive performance, SAGE's results can be more challenging to interpret, particularly when using conditional sampling. The choice of conditional sampler can significantly affect results, and there is limited empirical guidance on best practices. This vignette focuses on demonstrating the methods rather than making strong interpretive claims.

**The SAGE estimator** implemented here is what is referred to as the "permutation estimator" in other implementations.
It works by first building up `n_permutations` permutations of the feature vector and then successively evaluating prefixes of the sequence from left to right as coalitions to evaluated. 
If a task has features `(x1, x2, x3)`, one permutation could be `(x2, x1, x3)`, resulting in these coalitions to be evaluated: `(x2)`, `(x2, x1)`, and `(x2, x1, x3)`.
The empty coalition will always be evaluated, resulting in a total number of evaluations of `n_permutations * n_features + 1`.

## Demonstration with Correlated Features

To showcase the difference between Marginal and Conditional SAGE, we'll use the `sim_dgp_correlated()` function which creates a simple linear DGP with two correlated features.

**Model:**
$$(X_1, X_2)^T \sim \text{MVN}(0, \Sigma)$$

where $\Sigma$ is a 2×2 covariance matrix with 1 on the diagonal and correlation $r$ (default 0.5) on the off-diagonal.

$$X_3 \sim N(0,1), \quad X_4 \sim N(0,1)$$
$$Y = 2 \cdot X_1 + X_3 + \varepsilon$$

where $\varepsilon \sim N(0, 0.2^2)$.

**Data generating process:**

- `x1` has a direct effect on y (β=2.0)
- `x2` is correlated with x1 (r = 0.5) but has no direct effect on y
- `x3` is independent with a direct effect (β=1.0)
- `x4` is independent noise (β=0)

```{r data-setup}
set.seed(123)
task = sim_dgp_correlated(n = 1000, r = 0.5)

# Check the correlation structure
task_data = task$data()
correlation_matrix = cor(task_data[, c("x1", "x2", "x3", "x4")])
round(correlation_matrix, 3)
```

This DGP allows us to observe how the two SAGE variants handle correlated features with different roles in the data generating process.

Let's set up our learner and measure. We'll use a random forest and instantiate a resampling to ensure both methods see the same data:

```{r learner-setup}
learner = lrn("regr.ranger")
measure = msr("regr.mse")
resampling = rsmp("holdout")
resampling$instantiate(task)
```

## Marginal SAGE

Marginal SAGE marginalizes features independently by averaging predictions over a subset of `n_samples` observations drawn from the test dataset.
We use 15 permutations of the feature vector to build coalitions, resulting in 61 evaluated coalitions (`15 * 4 + 1`).

```{r marginal-sage}
# Create Marginal SAGE instance
marginal_sage = MarginalSAGE$new(
	task = task,
	learner = learner,
	measure = measure,
	resampling = resampling,
	n_permutations = 15L,
	n_samples = 100L
)

# Compute SAGE values
marginal_sage$compute(batch_size = 5000L)
```

Let's visualize the results:

```{r marginal-sage-plot}
#| echo: false
#| fig.alt: "Bar chart with features on x-axis and SAGE values on y-axis. Bars colored by feature type (causal vs noise)."
# Extract importance scores
marginal_results = marginal_sage$importance()
marginal_results$method = "Marginal SAGE"

# Create a factor with proper ordering
marginal_results$feature = factor(
	marginal_results$feature,
	levels = marginal_results$feature[order(marginal_results$importance, decreasing = TRUE)]
)

# Color features by type (causal vs noise)
marginal_results$feature_type = ifelse(
	marginal_results$feature %in% c("x1", "x3"),
	"Causal",
	"Noise"
)

# Create bar plot
ggplot(marginal_results, aes(x = feature, y = importance)) +
	geom_col(aes(fill = feature_type), alpha = 0.8) +
	scale_fill_manual(
		values = c("Causal" = "steelblue", "Noise" = "lightcoral"),
		name = "Feature type"
	) +
	labs(
		title = "Marginal SAGE Feature Importance",
		subtitle = "Correlated features: x1 and x2 are correlated (r = 0.5 from MVN)\nx2 has no effect on y",
		x = "Features",
		y = "SAGE Value"
	) +
	theme_minimal(base_size = 14)
```

We can also keep track of the SAGE value approximation across permutations:

```{r convergence-marginal}
#| fig.alt: "Line plot with permutation number on x-axis and SAGE value on y-axis. Multiple colored lines showing convergence trajectory for each feature."
marginal_sage$plot_convergence()
```

### Early Stopping Based on Convergence

SAGE supports early stopping to save computation time when the importance values have converged.
It is off by default; enable it with `early_stopping = TRUE`.
Convergence is detected by monitoring the standard error (SE) of the SAGE value estimates in the first resampling iteration.

SAGE normalizes the SE by the range of their values (max - min) to make convergence detection scale-invariant across different loss metrics. Convergence is detected when:

$$
\max_j \left(\frac{SE_j}{\max_i(\text{SAGE}_i) - \min_i(\text{SAGE}_i)}\right) < \text{threshold}
$$

The default threshold is `se_threshold = 0.025` (2.5%), meaning convergence occurs when the relative SE is below 2.5% of the importance range for all features, which is the criterion and default of the Python `sage` package.

With early stopping, `n_permutations` is an upper bound rather than a planned cost: if the criterion is not met within it, the values are returned with a warning.

You can customize convergence detection in `$compute()`:

```r
# More strict convergence (requires more permutations)
sage$compute(early_stopping = TRUE, se_threshold = 0.005, min_permutations = 5L)

# Disable early stopping to always run all permutations
sage$compute(early_stopping = FALSE)
```

After computation, `$budget` reports the requested and actually used effort, the implied number of evaluated coalitions, and whether the criterion was met:

```r
marginal_sage$budget
```

If a resampling with multiple iterations (i.e., not holdout) is supplied, the budget used by the first iteration is reused for all subsequent iterations to avoid some computational overhead.

### Exact SAGE for verification

When the number of features is small, you can sidestep coalition sampling entirely and compute the SAGE values *exactly* with `estimator = "exact"`, which enumerates all `2^p` coalitions:

```{r estimator-exact}
task_small = sim_dgp_correlated(n = 500) # four features
resampling_s = rsmp("holdout")
resampling_s$instantiate(task_small)

sage_exact = MarginalSAGE$new(
	task = task_small,
	learner = lrn("regr.ranger", num.trees = 50),
	measure = msr("regr.mse"),
	resampling = resampling_s,
	estimator = "exact",
	n_samples = 50L
)
sage_exact$compute()
sage_exact$importance()
sage_exact$budget
```

This is guarded by `max_features` (default 12), since the number of coalitions grows as `2^p`.
It carries no coalition-sampling error, so for `MarginalSAGE` it yields the exact Shapley values of the value function that the permutation estimator approximates, which makes it a convenient ground-truth reference when validating or comparing estimators.
The costs are directly comparable via the number of evaluated coalitions (`$budget$n_evals`): `1 + n_permutations * p` for the permutation estimator and `2^p` for exact enumeration.
When a permutation budget meets or exceeds the exact cost, `$compute()` points this out in a message (silenced by `xplain_opt(verbose = FALSE)`).
Note that "exact" refers to coalition sampling only: the *marginalization error* controlled by `n_samples` remains, and for `ConditionalSAGE` the value function is itself estimated by a sampler, so "exact" removes the coalition-sampling error but not the sampler's Monte Carlo error.

## Conditional SAGE

Conditional SAGE uses conditional sampling (via ARF by default) to marginalize features while preserving dependencies between the remaining features. This can provide different insights, especially when features are correlated.

```{r conditional-sage}
# Create Conditional SAGE instance using a conditional sampler
sampler_gaussian = ConditionalGaussianSampler$new(task)

conditional_sage = ConditionalSAGE$new(
	task = task,
	learner = learner,
	measure = measure,
	resampling = resampling,
	n_permutations = 15L,
	n_samples = 100L,
	sampler = sampler_gaussian
)

# Compute SAGE values
conditional_sage$compute(batch_size = 5000L)
```

Let's visualize the conditional SAGE results:

```{r conditional-sage-plot}
#| echo: false
#| fig.alt: "Bar chart with features on x-axis and SAGE values on y-axis. Bars colored by feature type (causal vs noise)."

# Extract importance scores
conditional_results = conditional_sage$importance()
conditional_results$method = "Conditional SAGE"

# Create a factor with proper ordering
conditional_results$feature = factor(
	conditional_results$feature,
	levels = conditional_results$feature[order(conditional_results$importance, decreasing = TRUE)]
)

# Color features by type (causal vs noise)
conditional_results$feature_type = ifelse(
	conditional_results$feature %in% c("x1", "x3"),
	"Causal",
	"Noise"
)

# Create bar plot
ggplot(conditional_results, aes(x = feature, y = importance)) +
	geom_col(aes(fill = feature_type), alpha = 0.8) +
	scale_fill_manual(
		values = c("Causal" = "steelblue", "Noise" = "lightcoral"),
		name = "Feature type"
	) +
	labs(
		title = "Conditional SAGE Feature Importance",
		subtitle = "Conditional sampling should reduce importance of redundant correlated features",
		x = "Features",
		y = "SAGE Value"
	) +
	theme_minimal(base_size = 14) +
	theme(legend.position = "bottom")
```


```{r convergence-conditional}
#| fig.alt: "Line plot with permutation number on x-axis and SAGE value on y-axis. Multiple colored lines showing convergence trajectory for each feature."
conditional_sage$plot_convergence()
```

## Comparison of Methods

Let's compare the two SAGE methods side by side:

```{r comparison}
#| echo: false
#| fig.alt: "Faceted grouped bar chart with two panels (Causal and Noise). Each panel shows features on x-axis with grouped bars for Marginal SAGE (blue) and Conditional SAGE (green)."

# Ensure both datasets have feature_type for consistency
marginal_results$feature_type = ifelse(
	marginal_results$feature %in% c("x1", "x3"),
	"Causal",
	"Noise"
)

# Combine results
combined_results = rbind(marginal_results, conditional_results)

# Create comparison plot
ggplot(combined_results, aes(x = feature, y = importance, fill = method)) +
	facet_wrap(~feature_type, scales = "free_x") +
	geom_col(position = "dodge", alpha = 0.8) +
	scale_fill_manual(values = c("Marginal SAGE" = "steelblue", "Conditional SAGE" = "darkgreen")) +
	labs(
		title = "Marginal vs Conditional SAGE Comparison",
		x = "Features",
		y = "SAGE Value",
		fill = "Method"
	) +
	theme_minimal(base_size = 14) +
	theme(legend.position = "bottom")
```

### Methodological Notes

The difference between the two methods:

- **MarginalSAGE**: Marginalizes all out-of-coalition features simultaneously by sampling from the marginal distribution, but does not account for conditional dependencies between in-coalition and out-of-coalition features.

- **ConditionalSAGE**: Uses conditional sampling to "marginalize" out-of-coalition features while preserving the conditional dependency structure between in-coalition and out-of-coalition features.

The interpretation of SAGE values, particularly for ConditionalSAGE, can be affected by the specific conditional sampler used and the nature of feature dependencies in the data.
The choice of sampler can significantly affect results, and there is currently limited empirical guidance on best practices for different settings.

## Comparison with PFI and CFI

For reference, let's compare SAGE methods with the analogous PFI and CFI methods on the same data:

```{r pfi-cfi-comparison}
# Quick PFI and CFI comparison for context
pfi = PFI$new(task, learner, measure)
cfi = CFI$new(task, learner, measure, sampler = sampler_gaussian)

pfi$compute()
cfi$compute()
pfi_results = pfi$importance()
cfi_results = cfi$importance()

# Create comparison data frame
method_comparison = data.frame(
	feature = rep(c("x1", "x2", "x3", "x4"), 4),
	importance = c(
		pfi_results$importance,
		cfi_results$importance,
		marginal_results$importance,
		conditional_results$importance
	),
	method = rep(c("PFI", "CFI", "Marginal SAGE", "Conditional SAGE"), each = 4),
	approach = rep(c("Marginal", "Conditional", "Marginal", "Conditional"), each = 4)
)

# Create comparison plot
#| fig.alt: "Grouped bar chart with features on x-axis and importance on y-axis. Four colored bars per feature for PFI, CFI, Marginal SAGE, and Conditional SAGE."
ggplot(method_comparison, aes(x = feature, y = importance, fill = method)) +
	geom_col(position = "dodge", alpha = 0.8) +
	scale_fill_manual(
		values = c(
			"PFI" = "lightblue",
			"CFI" = "blue",
			"Marginal SAGE" = "lightcoral",
			"Conditional SAGE" = "darkred"
		)
	) +
	labs(
		title = "Comparison: PFI/CFI vs Marginal/Conditional SAGE",
		subtitle = "Comparing perturbation-based and Shapley-based importance methods",
		x = "Features",
		y = "Importance Value",
		fill = "Method"
	) +
	theme_minimal(base_size = 14) +
	theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

While both PFI/CFI and MarginalSAGE/ConditionalSAGE distinguish between marginal and conditional approaches, these method families measure fundamentally different quantities. PFI and CFI measure the drop in predictive performance when features are perturbed, making their interpretation in terms of prediction loss relatively straightforward. SAGE methods measure each feature's contribution to overall performance through Shapley value decomposition, which involves a different theoretical framework. The results shown here demonstrate the methods on the same data, but direct comparisons of the numerical values should be made with these methodological differences in mind.
